### 2014-05-14 doc/ApTranscriptome_manuscript.Rmd
Edits to ms

**Authors**: John Stanton-Geddes, Andrew Nguyen, Rob Dunn, Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

@@

@@

We quantified expression of each gene using the program `Sailfish` [@patro2013] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

@@

@@

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 96,253 contigs and a total of 105,536 genes. For all genes, the total length was 63 million bp with an N~50~ length of 895 bp and a mean contig size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. Of these, 5,787 genes have best hits to putative genes in the genome sequence of the ant species *Solenopsis invicta* and 47% have best hits to genes from the class Insecta.

@@

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. carolinensis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")

mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
```

@@

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*A. carolinensis*) and colder (*A. picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression was greater in the more southern *A. carolinensis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. carolinensis* we also found that constitutive expression was greater in *A. carolinensis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range) for the *Intermediate* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. carolinensis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

>>>>

**Authors**: John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn, Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental Information for exact parameters). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmeider2011] with the provided bacteria, virus, archae and human databases of contaminants.

@@

We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

@@

We quantified expression of each gene using the program `Sailfish` [@patro2013] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of gene expression. We included the contaminant genes identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other genes, but removed these from further analyses. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

@@

Finally, to test whether the temperature at which thremally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression. We grouped the *High* genes along with the high temperature end of the *Bimodal* genes, and did the same for *Low* genes. We then tested if the temperature at which the *High* or *Low* group differed among the two species.

@@

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 96,253 contigs and a total of 105,536 genes. For all genes, the total length was 63 million bp with an N~50~ length of 895 bp and a mean contig size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts. Of these, 47% have best hits to genes from the class Insecta.

@@

We identified 22,399 genes with overall model fits of P < 0.05, retaining 10,597 after correcting for FDR. Of these transcripts, 9,156 significantly change expression with temperature including 6,908 that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. carolinensis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
```

```{r expression_type_plot, echo=FALSE}
mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
```

@@

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*A. carolinensis*) and colder (*A. picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater expression near optimal temperatures of 19°C in the warmer-climate species, and vice versa. Partially consistent with our prediction, we found that optimal-temperature expression was greater in the more southern *A. carolinensis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. carolinensis* we also found that optimal-temperature expression was greater in *A. carolinensis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`). Moreover, we also found that for genes with *Intermediate* expression in either colony, optimal-temperature expression was greater in *A. carolinensis*. The overall conclusion from these analyses is that for thermally-responsive genes, the more southern species, *A. carolinensis* has greater optimal-temperature expression.

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range for the *Intermediate* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. carolinensis* had a transcriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We also explored the extent to which the temperature at which thermally-responsive genes are activated differs between species. For genes that are upregulated at *High* temperatures, including both *High* and *Bimodal* genes, we found that ... In contrast, for genes that are upregulated at *Low* temperatures, including both *Low* and *Bimodal* genes, we found that...
